---
title: "ProteinBatcher Visualizer"
author: "Almirall S.A. & CRG/UPF Proteomics Unit"
format:
    dashboard:
        theme: flatly
        css: styles.css
        source: embed
        scrolling: true
server: shiny
runtime: shiny
---

```{r setup}
#| context: server-start

if (!require("ProteinBatcher")) {
    remotes::install_github("DataScienceRD-Almirall/ProteinBatcher")
}
library(ProteinBatcher)
library(SummarizedExperiment)
library(S4Vectors)
library(shiny)
library(bslib)
library(plotly)
library(DT)
library(ggplot2)
library(dplyr)

shiny::addResourcePath(
    "proteinbatcher-extdata",
    system.file("extdata", package = "ProteinBatcher")
)

options(shiny.maxRequestSize = 100 * 1024^2)

# ── Almirall colour palette (shared across all plots) ─────────────────────────
almirall_cols <- c(
    "#002855", "#008C94", "#00F0BE", "#8700D4",
    "#00BFA0", "#005A6E", "#fd7e14", "#dc3545",
    "#6f42c1", "#20c997", "#0dcaf0", "#ffc107"
)

condition_palette <- function(n) colorRampPalette(almirall_cols)(n)

# ── ggplot2 theme aligned with Almirall brand ──────────────────────────────────
theme_almirall <- function(base_size = 12) {
    ggplot2::theme_minimal(base_size = base_size, base_family = "sans") +
        ggplot2::theme(
            plot.title         = ggplot2::element_text(face = "bold", size = base_size + 2,
                                                       color = "#002855"),
            plot.subtitle      = ggplot2::element_text(size = base_size - 1, color = "#6c757d"),
            axis.text          = ggplot2::element_text(color = "#333"),
            axis.title         = ggplot2::element_text(color = "#333", size = base_size - 1),
            panel.grid.major   = ggplot2::element_line(color = "#e9ecef", linewidth = 0.4),
            panel.grid.minor   = ggplot2::element_blank(),
            panel.grid.major.x = ggplot2::element_blank(),
            legend.position    = "bottom",
            legend.title       = ggplot2::element_text(face = "bold", size = base_size - 1),
            legend.text        = ggplot2::element_text(size = base_size - 2),
            strip.text         = ggplot2::element_text(face = "bold", color = "#002855")
        )
}

# ── Contrast file parser ───────────────────────────────────────────────────────
parse_contrasts_file <- function(path) {
    lines <- readLines(path, warn = FALSE)
    lines <- trimws(lines)
    lines <- lines[!grepl("^#", lines) & nchar(lines) > 0]

    sections    <- list()
    current_sec <- NULL

    for (ln in lines) {
        if (grepl("^\\[.+\\]$", ln)) {
            current_sec             <- gsub("\\[|\\]", "", ln)
            sections[[current_sec]] <- character(0)
        } else if (!is.null(current_sec)) {
            sections[[current_sec]] <- c(sections[[current_sec]], ln)
        }
    }

    design <- list()
    for (item in sections[["design"]]) {
        idx <- regexpr("=", item)
        if (idx > 0) {
            key           <- trimws(substr(item, 1, idx - 1))
            val           <- trimws(substr(item, idx + 1, nchar(item)))
            design[[key]] <- val
        }
    }

    if (is.null(design[["formula"]]))
        stop("Missing 'formula' in [design] section.")
    if (is.null(design[["reference_condition"]]))
        stop("Missing 'reference_condition' in [design] section.")
    if (is.null(sections[["tests"]]) || length(sections[["tests"]]) == 0)
        stop("No tests listed in [tests] section.")

    tests_interaction <-
        if (!is.null(sections[["interaction_tests"]]) &&
            length(sections[["interaction_tests"]]) > 0)
            sections[["interaction_tests"]]
        else
            "NA"

    list(
        formula             = as.formula(design[["formula"]]),
        reference_condition = design[["reference_condition"]],
        tests               = sections[["tests"]],
        tests_interaction   = tests_interaction
    )
}
```

# Getting Started

## Uploading {.sidebar}

```{r sidebar-ui}
accordion(
    accordion_panel(
        title = "Input files",
        icon  = icon("upload"),

        tags$div(style = "margin-bottom:16px;",
            fileInput("annotation",
                tags$span("Annotation file", tags$code("(.tsv)")),
                accept = ".tsv"),
            tags$div(class = "example-links",
                icon("circle-info", style = "color:#6c757d;font-size:.8em;"), " Example: ",
                downloadLink("example_annotation", "annotation_HaCaT.tsv"))
        ),
        tags$div(
            fileInput("counts",
                tags$span("DIA-NN protein matrix", tags$code("(.pg_matrix.tsv)")),
                accept = ".tsv"),
            tags$div(class = "example-links",
                icon("circle-info", style = "color:#6c757d;font-size:.8em;"), " Example: ",
                downloadLink("example_counts", "HaCaT_pg_matrix.tsv"))
        )
    ),

    accordion_panel(
        title = "Contrasts",
        icon  = icon("shuffle"),

        fileInput("contrasts_file",
            tags$span("Contrasts file", tags$code("(.txt)")),
            accept = ".txt"),
        tags$div(class = "example-links",
            icon("circle-info", style = "color:#6c757d;font-size:.8em;"), " Example: ",
            downloadLink("example_contrasts", "example_contrasts.txt")),
        uiOutput("contrasts_preview")
    ),

    accordion_panel(
        title = "Parameters",
        icon  = icon("sliders-h"),

        sliderInput("percent_missing", "Maximum missingness per protein (%)",
                    min = 0, max = 100, value = 50, step = 5),
        sliderInput("threshold", "Mean-imputation threshold per condition (%)",
                    min = 0, max = 100, value = 50, step = 5),
        textInput("experiment", "Experiment name",
                  value = "MyExperiment", placeholder = "e.g. HaCaT"),
        selectInput("ldv_source", "LDV imputation source",
                    choices = c("per-condition", "global"), selected = "per-condition"),
        numericInput("alpha", "Adjusted p-value threshold (α)",
                     value = 0.05, min = 0, max = 1, step = 0.01),
        numericInput("lfc", "Log\u2082 fold-change threshold",
                     value = 1, min = 0, step = 0.1),
        checkboxInput("paired",           "Paired design",                       value = FALSE),
        checkboxInput("block_effect",     "Block effect (duplicateCorrelation)", value = FALSE),
        checkboxInput("test_interaction", "Test interaction contrasts",           value = TRUE)
    ),

    accordion_panel(
        title = "Run analysis",
        icon  = icon("play"),

        actionButton("run_analysis",
                     label = tagList(icon("flask"), " Run ProteinBatcher"),
                     class = "btn-primary w-100"),
        tags$div(style = "margin-top:14px;"),
        uiOutput("run_status_ui")
    ),

    open = 1
)
```

```{r server}
#| context: server

# ── Shared reactive state (all tabs read from rv) ──────────────────────────────
rv <- reactiveValues(
    contrasts         = NULL,
    contrasts_error   = NULL,
    se_raw            = NULL,
    se_filt           = NULL,
    removed           = NULL,
    se_imp            = NULL,
    se_limma          = NULL,
    effects           = NULL,
    tests             = NULL,
    tests_interaction = NULL,
    status            = "idle",
    error_msg         = NULL
)

# ── Example file downloads ─────────────────────────────────────────────────────
output$example_annotation <- downloadHandler(
    filename    = "annotation_HaCaT.tsv",
    content     = function(f) file.copy(
        system.file("extdata/annotation_HaCaT.tsv", package = "ProteinBatcher"), f),
    contentType = "text/tab-separated-values"
)
output$example_counts <- downloadHandler(
    filename    = "example_pg_matrix.tsv",
    content     = function(f) file.copy(
        system.file(
            "extdata/2024MK017_HaCaT_Stimulation_1to24_Astral_report.pg_matrix.tsv",
            package = "ProteinBatcher"), f),
    contentType = "text/tab-separated-values"
)
output$example_contrasts <- downloadHandler(
    filename    = "example_contrasts.txt",
    content     = function(f) file.copy(
        system.file("extdata/example_contrasts.txt", package = "ProteinBatcher"), f),
    contentType = "text/plain"
)

# ── Parse contrasts on upload ──────────────────────────────────────────────────
observeEvent(input$contrasts_file, {
    rv$contrasts       <- NULL
    rv$contrasts_error <- NULL
    result <- tryCatch(parse_contrasts_file(input$contrasts_file$datapath),
                       error = function(e) e)
    if (inherits(result, "error")) rv$contrasts_error <- conditionMessage(result)
    else                           rv$contrasts        <- result
})

# ── Contrasts sidebar preview ──────────────────────────────────────────────────
output$contrasts_preview <- renderUI({
    if (!is.null(rv$contrasts_error))
        return(tags$div(class = "alert alert-danger p-2 mt-2",
                        style = "font-size:.8em;",
                        icon("triangle-exclamation"), " ", rv$contrasts_error))
    req(rv$contrasts)
    c       <- rv$contrasts
    has_int <- !identical(c$tests_interaction, "NA")

    tags$div(class = "mt-2 p-2 border rounded bg-light",
             style = "font-size:.82em;line-height:1.6;",
        tags$div(icon("flask", style="color:#0d6efd;"),
                 tags$b(" Formula: "), tags$code(deparse(c$formula))),
        tags$div(icon("bookmark", style="color:#0d6efd;"),
                 tags$b(" Reference: "), tags$code(c$reference_condition)),
        tags$hr(style="margin:4px 0;"),
        tags$div(tags$b(icon("vials"), " Tests:")),
        tags$ul(class="mb-0 ps-3",
                lapply(c$tests, function(t) tags$li(tags$code(t)))),
        if (has_int) tagList(
            tags$div(class="mt-1", tags$b(icon("arrows-left-right"), " Interactions:")),
            tags$ul(class="mb-0 ps-3",
                    lapply(c$tests_interaction, function(t) tags$li(tags$code(t))))
        )
    )
})

# ── Run-button status badge ────────────────────────────────────────────────────
output$run_status_ui <- renderUI({
    switch(rv$status,
        idle    = tags$div(class="text-muted", style="font-size:.85em;",
                           icon("circle-info"), " Upload files and press Run."),
        running = tags$div(class="text-warning", style="font-size:.85em;",
                           icon("spinner", class="fa-spin"), " Pipeline running\u2026"),
        done    = tags$div(style="font-size:.85em;",
                     tags$div(class="text-success fw-bold",
                              icon("circle-check"), " Analysis complete"),
                     tags$div(class="text-muted mt-1",
                         sprintf("%d \u2192 %d proteins | %d contrast(s)",
                                 nrow(rv$se_raw), nrow(rv$se_filt), length(rv$tests)))),
        error   = tags$div(
                     tags$div(class="text-danger fw-bold",
                              icon("triangle-exclamation"), " Error"),
                     tags$div(class="text-danger mt-1 font-monospace",
                              style="font-size:.78em;word-break:break-word;",
                              rv$error_msg))
    )
})

# ── Run pipeline ───────────────────────────────────────────────────────────────
observeEvent(input$run_analysis, {
    if (is.null(input$annotation)) {
        rv$status <- "error"
        rv$error_msg <- "Please upload an annotation file (.tsv)."
        return()
    }
    if (is.null(input$counts)) {
        rv$status <- "error"
        rv$error_msg <- "Please upload a DIA-NN protein matrix (.pg_matrix.tsv)."
        return()
    }
    if (is.null(rv$contrasts)) {
        rv$status <- "error"
        rv$error_msg <- if (!is.null(rv$contrasts_error))
            paste("Contrasts file error:", rv$contrasts_error)
        else "Please upload a valid contrasts file (.txt)."
        return()
    }
    if (trimws(input$experiment) == "") {
        rv$status <- "error"
        rv$error_msg <- "Please enter an experiment name."
        return()
    }

    rv$status <- "running"; rv$error_msg <- NULL

    tryCatch({
        ct        <- rv$contrasts
        tests_int <- if (input$test_interaction && !identical(ct$tests_interaction, "NA"))
            ct$tests_interaction else "NA"

        withProgress(message = "ProteinBatcher pipeline", value = 0, {
            incProgress(0.10, detail = "Importing data\u2026")
            se_raw <- ProteinBatcher:::make_se_from_files(
                quant_table_path = input$counts$datapath,
                exp_anno_path    = input$annotation$datapath,
                type = "DIA", level = "protein")

            incProgress(0.15, detail = "Filtering by missingness\u2026")
            filt <- filter_se_missing(se_raw, percentage = input$percent_missing)

            incProgress(0.20, detail = "Imputing missing values\u2026")
            se_imp <- impute_se(filt$se_filt,
                                threshold  = input$threshold / 100,
                                ldv_source = input$ldv_source)

            incProgress(0.30, detail = "Fitting limma models\u2026")
            se_limma <- test_limma_customized(
                se               = se_imp,
                test             = ct$tests,
                test_interaction = tests_int,
                design_formula   = ct$formula,
                ref_condition    = ct$reference_condition,
                paired           = input$paired,
                block_effect     = input$block_effect)

            incProgress(0.20, detail = "Organising effects\u2026")
            effects <- ProteinBatcher:::.pp_organize_effects(
                se_limma          = se_limma,
                tests             = ct$tests,
                tests_interaction = tests_int,
                alpha             = input$alpha)

            rv$se_raw            <- se_raw
            rv$se_filt           <- filt$se_filt
            rv$removed           <- filt$removed
            rv$se_imp            <- se_imp
            rv$se_limma          <- se_limma
            rv$effects           <- effects
            rv$tests             <- ct$tests
            rv$tests_interaction <- tests_int
            rv$status            <- "done"
            incProgress(0.05, detail = "Done!")
        })
    }, error = function(e) {
        rv$status <- "error"; rv$error_msg <- conditionMessage(e)
    })
})

# ── Getting Started: pipeline summary card ─────────────────────────────────────
output$pipeline_summary_ui <- renderUI({
    if (rv$status == "idle") return(tags$div(
        class="text-center text-muted py-4", style="font-size:.95em;",
        icon("arrow-left", style="font-size:1.4em;"), tags$br(), tags$br(),
        "Upload your files, define contrasts,", tags$br(),
        "then click ", tags$b("Run ProteinBatcher"), "."))

    if (rv$status == "running") return(tags$div(
        class="text-center text-warning py-4",
        icon("spinner", class="fa-spin", style="font-size:2em;"),
        tags$br(), tags$br(), "Pipeline running, please wait\u2026"))

    if (rv$status == "error") return(tags$div(
        class="alert alert-danger",
        icon("triangle-exclamation"), tags$b(" Pipeline error"),
        tags$hr(style="margin:6px 0;"), tags$code(rv$error_msg)))

    n_raw    <- nrow(rv$se_raw);  n_filt  <- nrow(rv$se_filt)
    pct_kept <- round(100 * n_filt / n_raw, 1)
    n_imp    <- sum(rowData(rv$se_imp)$imputed, na.rm = TRUE)
    pct_imp  <- round(100 * n_imp / n_filt, 1)
    has_int  <- !identical(rv$tests_interaction, "NA")

    mc <- function(v, lbl, col) tags$div(
        class="border rounded p-2 text-center", style="flex:1;",
        tags$div(style=paste0("font-size:1.6em;font-weight:700;color:",col,";"), v),
        tags$div(class="text-muted", style="font-size:.78em;", lbl))

    tags$div(
        tags$div(style="display:flex;gap:8px;flex-wrap:wrap;margin-bottom:12px;",
            mc(ncol(rv$se_raw), "Samples", "#0d6efd"),
            mc(n_raw,           "Proteins (raw)", "#6c757d"),
            mc(paste0(n_filt," (",pct_kept,"%)"), "After filter", "#198754"),
            mc(paste0(n_imp, " (",pct_imp, "%)"), "Imputed",      "#fd7e14")),
        tags$hr(style="margin:8px 0;"),
        tags$div(tags$b(icon("vials"), " Main contrasts:"),
            tags$ul(class="mb-1 mt-1 ps-3", style="font-size:.88em;",
                lapply(rv$tests, function(t)
                    tags$li(icon("check", style="color:#198754;"), " ", tags$code(t))))),
        if (has_int) tags$div(
            tags$b(icon("arrows-left-right"), " Interaction contrasts:"),
            tags$ul(class="mb-0 mt-1 ps-3", style="font-size:.88em;",
                lapply(rv$tests_interaction, function(t)
                    tags$li(icon("check", style="color:#0dcaf0;"), " ", tags$code(t))))),
        tags$hr(style="margin:8px 0;"),
        tags$div(class="text-success text-center", style="font-size:.9em;font-weight:600;",
                 icon("circle-check"), " Navigate to the other tabs to explore your results.")
    )
})

# ════════════════════════════════════════════════════════════════════════════════
# DATA INSPECTION — all server logic
# ════════════════════════════════════════════════════════════════════════════════

# ── Update sidebar selectors when pipeline completes ──────────────────────────
observe({
    req(rv$status == "done", rv$tests)
    updateSelectInput(session, "di_contrast",
                      choices = rv$tests, selected = rv$tests[1])
})

observe({
    req(rv$status == "done", rv$se_imp)
    rd <- as.data.frame(rowData(rv$se_imp))
    gene_choices <- if ("Genes" %in% colnames(rd)) {
        setNames(rownames(rd),
                 ifelse(!is.na(rd$Genes) & rd$Genes != "",
                        paste0(rd$Genes, "  [", rownames(rd), "]"),
                        rownames(rd)))
    } else {
        setNames(rownames(rd), rownames(rd))
    }
    # server = FALSE -> all choices loaded client-side: instant fuzzy search
    updateSelectizeInput(session, "di_gene",
                         choices = gene_choices, server = FALSE)
})

observe({
    req(rv$status == "done", rv$se_imp)
    cd_cols   <- colnames(as.data.frame(colData(rv$se_imp)))
    meta_cols <- intersect(c("condition", "batch", "donor_id", "replicate"), cd_cols)
    if (length(meta_cols) == 0) return()
    updateSelectInput(session, "di_color_by",
                      choices = meta_cols, selected = meta_cols[1])
    shape_opts <- c("None", setdiff(meta_cols, meta_cols[1]))
    updateSelectInput(session, "di_shape_by",
                      choices  = shape_opts,
                      selected = if (length(shape_opts) > 1) shape_opts[2] else "None")
    updateSelectInput(session, "di_dot_color",
                      choices = meta_cols, selected = meta_cols[1])
    updateSelectInput(session, "di_dot_shape",
                      choices  = c("None", meta_cols),
                      selected = if ("batch" %in% meta_cols) "batch" else "None")
})

# ── PCA reactive (cached — only recomputes when se_imp changes) ────────────────
pca_data <- reactive({
    req(rv$se_imp)
    mat          <- assay(rv$se_imp)
    mat_complete <- mat[complete.cases(mat), ]
    n_prot       <- nrow(mat_complete)
    pca_obj      <- prcomp(t(mat_complete), scale. = TRUE, center = TRUE)
    var_exp      <- round(100 * pca_obj$sdev^2 / sum(pca_obj$sdev^2), 1)
    n_pc         <- min(10, ncol(pca_obj$x))
    pca_df       <- as.data.frame(pca_obj$x[, seq_len(n_pc), drop = FALSE])
    pca_df$sample_name <- rownames(pca_df)

    cd <- as.data.frame(colData(rv$se_imp))
    for (col in c("condition", "batch", "donor_id", "replicate"))
        if (col %in% colnames(cd))
            pca_df[[col]] <- cd[pca_df$sample_name, col]

    list(df = pca_df, var_exp = var_exp, n_prot = n_prot, n_pc = n_pc)
})

# Update PC axis choices after PCA runs
observe({
    req(pca_data())
    pc_choices <- paste0("PC", seq_len(pca_data()$n_pc))
    updateSelectInput(session, "di_pc_x", choices = pc_choices, selected = "PC1")
    updateSelectInput(session, "di_pc_y", choices = pc_choices, selected = "PC2")
})

# ── PCA scatter plot ───────────────────────────────────────────────────────────
output$di_pca_plot <- renderPlotly({
    req(rv$status == "done")
    pca     <- pca_data()
    df      <- pca$df
    pc_x    <- if (!is.null(input$di_pc_x) && input$di_pc_x %in% colnames(df)) input$di_pc_x else "PC1"
    pc_y    <- if (!is.null(input$di_pc_y) && input$di_pc_y %in% colnames(df)) input$di_pc_y else "PC2"
    col_var <- if (!is.null(input$di_color_by) && input$di_color_by %in% colnames(df)) input$di_color_by else "condition"
    shp_var <- if (!is.null(input$di_shape_by) && input$di_shape_by != "None" &&
                   input$di_shape_by %in% colnames(df)) input$di_shape_by else NULL

    conds     <- as.character(unique(df[[col_var]]))
    color_map <- setNames(condition_palette(length(conds)), conds)

    pc_x_i <- suppressWarnings(as.integer(sub("PC", "", pc_x)))
    pc_y_i <- suppressWarnings(as.integer(sub("PC", "", pc_y)))
    var_x  <- if (!is.na(pc_x_i) && pc_x_i <= length(pca$var_exp)) pca$var_exp[pc_x_i] else "?"
    var_y  <- if (!is.na(pc_y_i) && pc_y_i <= length(pca$var_exp)) pca$var_exp[pc_y_i] else "?"

    hover_cols    <- intersect(c("condition", "batch", "donor_id", "replicate"), colnames(df))
    df$hover_text <- apply(df[, c("sample_name", hover_cols), drop = FALSE], 1, function(row) {
        meta <- paste(hover_cols, row[hover_cols], sep = ": ", collapse = "<br>")
        paste0("<b>", row["sample_name"], "</b><br>", meta)
    })

    # Add symbol column for shape mapping
    if (!is.null(shp_var)) {
        df$shp_col <- factor(df[[shp_var]])
    }

    p <- plot_ly(
        data        = df,
        x           = as.formula(paste0("~`", pc_x, "`")),
        y           = as.formula(paste0("~`", pc_y, "`")),
        color       = as.formula(paste0("~factor(`", col_var, "`)")) ,
        colors      = color_map,
        symbol      = if (!is.null(shp_var)) as.formula(paste0("~shp_col")) else NULL,
        text        = ~hover_text,
        hoverinfo   = "text",
        type        = "scatter",
        mode        = "markers",
        marker      = list(size = 11, opacity = 0.88,
                           line = list(color = "white", width = 1))
    ) %>%
        layout(
            title = list(
                text = paste0("<b>PCA</b>  <span style='font-size:12px;color:#6c757d;'>",
                              pca$n_prot, " proteins, ", ncol(rv$se_imp), " samples</span>"),
                font = list(family = "Segoe UI", size = 15, color = "#002855"),
                x    = 0.01),
            xaxis  = list(title      = paste0(pc_x, "  (", var_x, "%)"),
                          gridcolor  = "#e9ecef", zerolinecolor = "#ced4da", zeroline = TRUE),
            yaxis  = list(title      = paste0(pc_y, "  (", var_y, "%)"),
                          gridcolor  = "#e9ecef", zerolinecolor = "#ced4da", zeroline = TRUE),
            legend = list(title       = list(text = paste0("<b>", col_var, "</b>")),
                          orientation = "h", y = -0.22, font = list(size = 11)),
            paper_bgcolor = "white", plot_bgcolor = "white",
            margin = list(t = 75, b = 80)
        ) %>%
        config(displayModeBar = TRUE,
               modeBarButtonsToRemove = c("lasso2d", "select2d", "autoScale2d"),
               toImageButtonOptions   = list(format = "png", filename = "PCA"))
})

# ── Scree plot ─────────────────────────────────────────────────────────────────
output$di_scree_plot <- renderPlotly({
    req(rv$status == "done")
    pca    <- pca_data()
    n_show <- min(pca$n_pc, 10)
    df     <- data.frame(
        PC     = factor(paste0("PC", seq_len(n_show)),
                        levels = paste0("PC", seq_len(n_show))),
        Var    = pca$var_exp[seq_len(n_show)]
    )
    df$cumVar <- cumsum(df$Var)

    plot_ly(df) %>%
        add_bars(x = ~PC, y = ~Var,
                 marker            = list(color = "#008C94",
                                          line  = list(color = "#005A6E", width = 1)),
                 name              = "% Variance",
                 hovertemplate     = "<b>%{x}</b><br>Variance: %{y:.1f}%<extra></extra>") %>%
        add_lines(x = ~PC, y = ~cumVar,
                  line          = list(color = "#8700D4", width = 2, dash = "dot"),
                  name          = "Cumulative",
                  yaxis         = "y2",
                  hovertemplate = "<b>%{x}</b><br>Cumulative: %{y:.1f}%<extra></extra>") %>%
        layout(
            title  = list(text = "<b>Variance explained</b>",
                          font = list(family = "Segoe UI", size = 13, color = "#002855"),
                          x = 0.01),
            xaxis  = list(title = NULL, tickfont = list(size = 10)),
            yaxis  = list(title = "% Variance", gridcolor = "#e9ecef", ticksuffix = "%"),
            yaxis2 = list(title = "Cumulative %", overlaying = "y", side = "right",
                          range = c(0, 105), gridcolor = "transparent", ticksuffix = "%"),
            legend = list(orientation = "h", y = -0.3, font = list(size = 10)),
            paper_bgcolor = "white", plot_bgcolor = "white",
            margin = list(t = 40, b = 55)
        ) %>%
        config(displayModeBar = FALSE)
})

# ── Metric cards row ───────────────────────────────────────────────────────────
output$di_metrics_ui <- renderUI({
    req(rv$status == "done", rv$se_limma)

    ct     <- req(input$di_contrast)
    rd     <- as.data.frame(rowData(rv$se_limma))
    padj_c <- paste0("p.adj_",  ct)
    lfc_c  <- paste0("log2FC_", ct)

    n_up <- n_down <- NA_integer_
    if (padj_c %in% colnames(rd) && lfc_c %in% colnames(rd)) {
        n_up   <- sum(rd[[padj_c]] <= input$alpha & rd[[lfc_c]] >=  input$lfc, na.rm = TRUE)
        n_down <- sum(rd[[padj_c]] <= input$alpha & rd[[lfc_c]] <= -input$lfc, na.rm = TRUE)
    }

    mk <- function(val, lbl, icon_name, bg, tc = "white") tags$div(
        class = "rounded p-3",
        style = paste0("background:", bg, ";flex:1;min-width:110px;"),
        tags$div(style = paste0("font-size:1.8em;font-weight:800;color:", tc, ";"), val),
        tags$div(style = paste0("font-size:.75em;color:", tc, ";opacity:.9;margin-top:2px;"),
                 icon(icon_name), " ", lbl))

    tags$div(style = "display:flex;gap:10px;flex-wrap:wrap;padding:4px 0;",
        mk(ncol(rv$se_imp),                 "Samples",           "vial",       "#002855"),
        mk(nrow(rv$se_raw),                 "Proteins detected", "microscope", "#005A6E"),
        mk(nrow(rv$se_filt),                "Proteins analysed", "filter",     "#008C94"),
        mk(sum(rowData(rv$se_imp)$imputed, na.rm = TRUE),
                                            "Proteins imputed",  "fill-drip",  "#8700D4"),
        if (!is.na(n_up))
            mk(paste0("\u2191", n_up, " / \u2193", n_down),
               paste0("DE  ", sub("_vs_.*", "", ct), " vs ", sub(".*_vs_", "", ct)),
               "chart-bar", "#fd7e14")
    )
})

# ── Protein abundance explorer ────────────────────────────────────────────────
output$di_protein_plot <- renderPlotly({
    req(rv$status == "done", input$di_gene, nchar(input$di_gene) > 0)

    gene_id <- input$di_gene
    mat     <- assay(rv$se_imp)
    if (!gene_id %in% rownames(mat))
        return(plotly_empty() %>%
                   layout(title = "Gene not found \u2014 try another search term."))

    rd      <- as.data.frame(rowData(rv$se_imp))
    cd      <- as.data.frame(colData(rv$se_imp))
    imp_map <- metadata(rv$se_imp)$imputation_map

    gene_label <- if ("Genes" %in% colnames(rd) && !is.na(rd[gene_id, "Genes"]) &&
                      nchar(rd[gene_id, "Genes"]) > 0) rd[gene_id, "Genes"] else gene_id
    prot_name  <- if ("Protein.Names" %in% colnames(rd) && !is.na(rd[gene_id, "Protein.Names"]))
        rd[gene_id, "Protein.Names"] else ""

    plot_df <- data.frame(
        sample_name = colnames(mat),
        intensity   = as.numeric(mat[gene_id, ]),
        imputed     = as.logical(imp_map[gene_id, ] != "none"),
        stringsAsFactors = FALSE
    )
    for (col in c("condition", "batch", "donor_id", "replicate"))
        if (col %in% colnames(cd)) plot_df[[col]] <- cd[plot_df$sample_name, col]

    # Condition ordering: reference first
    ref_cond  <- rv$contrasts$reference_condition
    cond_lev  <- c(ref_cond, setdiff(unique(plot_df$condition), ref_cond))
    plot_df$condition <- factor(plot_df$condition, levels = cond_lev)

    col_var   <- if (!is.null(input$di_dot_color) && input$di_dot_color %in% colnames(plot_df))
        input$di_dot_color else "condition"
    cond_vals <- as.character(unique(plot_df[[col_var]]))
    color_map <- setNames(condition_palette(length(cond_vals)), cond_vals)

    # Hover
    hover_cols  <- intersect(c("batch", "donor_id", "replicate"), colnames(plot_df))
    plot_df$hover <- paste0(
        "<b>", plot_df$sample_name, "</b>",
        if (length(hover_cols) > 0)
            apply(plot_df[, hover_cols, drop = FALSE], 1, function(r)
                paste0("<br>", paste(hover_cols, r, sep = ": ", collapse = "<br>")))
        else "",
        "<br>Log\u2082 intensity: <b>", round(plot_df$intensity, 3), "</b>",
        ifelse(plot_df$imputed,
               "<br><span style='color:#fd7e14;'>\u26a0 imputed value</span>", "")
    )

    # Pre-compute ONE mean per condition (avoids ggplotly stat_summary bug)
    cond_means <- aggregate(intensity ~ condition, data = plot_df,
                            FUN = mean, na.rm = TRUE)
    cond_means$condition <- factor(cond_means$condition,
                                   levels = levels(plot_df$condition))
    cond_means$bar_hover <- paste0(
        "<b>Mean (", cond_means$condition, ")</b>: ",
        round(cond_means$intensity, 3)
    )

    gg <- ggplot2::ggplot(plot_df,
            ggplot2::aes(x = condition, y = intensity,
                         color = .data[[col_var]], text = hover)) +
        # ONE horizontal bar per condition = condition mean
        ggplot2::geom_errorbar(
            data = cond_means,
            ggplot2::aes(x = condition, ymin = intensity, ymax = intensity,
                         text = bar_hover),
            width = 0.44, linewidth = 2.4,
            color = "#002855", alpha = 0.85,
            inherit.aes = FALSE
        ) +
        # Observed points
        ggplot2::geom_jitter(
            data  = plot_df[!plot_df$imputed, ],
            width = 0.12, size = 3.2, alpha = 0.85, shape = 16
        ) +
        # Imputed points (open circles)
        ggplot2::geom_jitter(
            data   = plot_df[plot_df$imputed, ],
            width  = 0.12, size = 3.5, alpha = 0.55, shape = 1, stroke = 1.5
        ) +
        ggplot2::scale_color_manual(values = color_map, guide = "none") +
        ggplot2::labs(
            title    = gene_label,
            subtitle = if (nchar(prot_name) > 0) prot_name else NULL,
            x        = NULL,
            y        = "Protein Abundance (Log\u2082)"
        ) +
        theme_almirall() +
        ggplot2::theme(
            axis.text.x   = ggplot2::element_text(angle = 30, hjust = 1, size = 10),
            plot.title    = ggplot2::element_text(size = 14, face = "bold", color = "#002855"),
            plot.subtitle = ggplot2::element_text(size = 9, color = "#6c757d")
        )

    plotly::ggplotly(gg, tooltip = "text") %>%
        plotly::layout(margin = list(t = 85, b = 100)) %>%
        plotly::config(
            displayModeBar = TRUE,
            modeBarButtonsToRemove = c("lasso2d", "select2d", "autoScale2d"),
            toImageButtonOptions   = list(format = "png",
                                          filename = paste0(gene_label, "_abundance"))
        )
})

# ── Per-contrast DE summary table ─────────────────────────────────────────────
output$di_summary_table <- DT::renderDT({
    req(rv$status == "done", rv$se_limma)

    rd    <- as.data.frame(rowData(rv$se_limma))
    alpha <- input$alpha
    lfc   <- input$lfc

    rows <- lapply(rv$tests, function(ct) {
        padj_c <- paste0("p.adj_",  ct)
        lfc_c  <- paste0("log2FC_", ct)
        if (!(padj_c %in% colnames(rd) && lfc_c %in% colnames(rd)))
            return(NULL)

        n_tested <- sum(!is.na(rd[[padj_c]]))
        n_up     <- sum(rd[[padj_c]] <= alpha & rd[[lfc_c]] >=  lfc, na.rm = TRUE)
        n_down   <- sum(rd[[padj_c]] <= alpha & rd[[lfc_c]] <= -lfc, na.rm = TRUE)
        n_ns     <- n_tested - n_up - n_down

        imp_map  <- metadata(rv$se_imp)$imputation_map
        cd       <- as.data.frame(colData(rv$se_imp))
        parts    <- strsplit(ct, "_vs_")[[1]]
        g1 <- parts[1]; g2 <- parts[2]
        imp_g1 <- if (g1 %in% cd$condition)
            apply(imp_map[, cd$condition == g1, drop = FALSE], 1,
                  function(x) any(x != "none")) else rep(FALSE, nrow(imp_map))
        imp_g2 <- if (g2 %in% cd$condition)
            apply(imp_map[, cd$condition == g2, drop = FALSE], 1,
                  function(x) any(x != "none")) else rep(FALSE, nrow(imp_map))
        n_imp_any <- sum(imp_g1 | imp_g2, na.rm = TRUE)

        data.frame(
            Contrast = ct,
            Tested   = n_tested,
            Up       = n_up,
            Down     = n_down,
            NS       = n_ns,
            Imputed  = n_imp_any,
            stringsAsFactors = FALSE
        )
    })

    tbl         <- do.call(rbind, rows[!sapply(rows, is.null)])
    colnames(tbl) <- c("Contrast", "Tested", "\u2191 Up", "\u2193 Down", "NS", "\u22651 imputed")

    DT::datatable(tbl,
        rownames = FALSE,
        class    = "table table-sm table-hover",
        options  = list(dom = "t", ordering = FALSE, pageLength = 20)
    ) %>%
        DT::formatStyle("\u2191 Up",
            color = "white",
            backgroundColor = DT::styleInterval(0, c("transparent", "#dc3545")),
            fontWeight = "bold") %>%
        DT::formatStyle("\u2193 Down",
            color = "white",
            backgroundColor = DT::styleInterval(0, c("transparent", "#0d6efd")),
            fontWeight = "bold") %>%
        DT::formatStyle("\u22651 imputed",
            color = DT::styleInterval(0, c("inherit", "#fd7e14")),
            fontWeight = "bold") %>%
        DT::formatStyle("Contrast",
            fontWeight = "bold", color = "#002855")

}, server = FALSE)

# ════════════════════════════════════════════════════════════════════════════════
# IMPUTATION & FILTERING — server logic
# ════════════════════════════════════════════════════════════════════════════════

# ── Update contrast selector ──────────────────────────────────────────────────
observe({
    req(rv$status == "done", rv$tests)
    updateSelectInput(session, "imp_contrast",
                      choices = rv$tests, selected = rv$tests[1])
})

# ── Helper: classify imputation status per protein per contrast ────────────────
.imp_classify <- function(imp_map, cd, c1, c2) {
    get_st <- function(x, grp) {
        idx <- which(cd$condition == grp)
        if (!length(idx)) return("none")
        vals <- x[idx]
        if (any(vals == "ldv")) "ldv" else if (any(vals == "mean")) "mean" else "none"
    }
    apply(imp_map, 1, function(x) paste0(get_st(x, c1), "-", get_st(x, c2)))
}

# ── Before/after intensity data ────────────────────────────────────────────────
before_after_data <- reactive({
    req(rv$se_filt, rv$se_imp)
    expr_b <- as.numeric(assay(rv$se_filt))
    expr_a <- as.numeric(assay(rv$se_imp))
    expr_b <- expr_b[!is.na(expr_b)]
    n_na_b <- sum(is.na(as.numeric(assay(rv$se_filt))))
    data.frame(
        expression = c(expr_b, expr_a),
        dataset    = c(rep("Before", length(expr_b)), rep("After", length(expr_a))),
        stringsAsFactors = FALSE
    )
})

# ── FC + imputation status data per contrast ───────────────────────────────────
imp_fc_data <- reactive({
    req(rv$status == "done", input$imp_contrast, rv$se_limma)
    ct    <- input$imp_contrast
    parts <- strsplit(ct, "_vs_")[[1]]
    if (length(parts) != 2) return(NULL)
    c1 <- parts[1]; c2 <- parts[2]
    imp_map <- metadata(rv$se_imp)$imputation_map
    cd      <- as.data.frame(colData(rv$se_imp))
    prot_st <- .imp_classify(imp_map, cd, c1, c2)
    rd      <- as.data.frame(rowData(rv$se_limma))
    lfc_col  <- paste0("log2FC_", ct)
    padj_col <- paste0("p.adj_", ct)
    if (!lfc_col %in% colnames(rd)) return(NULL)
    df <- data.frame(
        protein_id = rownames(rd),
        gene       = if ("Genes" %in% colnames(rd)) rd$Genes else rownames(rd),
        log2FC     = rd[[lfc_col]],
        p_adj      = if (padj_col %in% colnames(rd)) rd[[padj_col]] else NA_real_,
        imp_status = prot_st,
        stringsAsFactors = FALSE
    )
    df[!is.na(df$log2FC), , drop = FALSE]
})

# ── Imputation type summary ────────────────────────────────────────────────────
imp_summary_data <- reactive({
    req(imp_fc_data())
    df    <- imp_fc_data()
    alpha <- input$alpha; lfc <- input$lfc
    cats  <- sort(unique(df$imp_status))
    do.call(rbind, lapply(cats, function(cat) {
        s <- df[df$imp_status == cat, ]
        data.frame(
            Category      = cat,
            N             = nrow(s),
            Up            = sum(!is.na(s$p_adj) & s$p_adj <= alpha & s$log2FC >=  lfc, na.rm=TRUE),
            Down          = sum(!is.na(s$p_adj) & s$p_adj <= alpha & s$log2FC <= -lfc, na.rm=TRUE),
            Median_log2FC = round(median(s$log2FC, na.rm=TRUE), 3),
            stringsAsFactors = FALSE
        )
    }))
})

# Update category checkboxes — use bindEvent on imp_contrast to avoid cycle
# with the density/boxplot renders that also depend on imp_fc_data()
observe({
    req(rv$status == "done", input$imp_contrast)
    # Compute categories independently (don't go through imp_fc_data() reactive
    # to break the timing cycle with the plot renders)
    ct    <- input$imp_contrast
    parts <- strsplit(ct, "_vs_")[[1]]
    if (length(parts) != 2) return()
    c1 <- parts[1]; c2 <- parts[2]
    imp_map <- metadata(rv$se_imp)$imputation_map
    cd      <- as.data.frame(colData(rv$se_imp))
    prot_st <- .imp_classify(imp_map, cd, c1, c2)
    cats    <- sort(unique(prot_st))
    updateCheckboxGroupInput(session, "imp_cats_filter",
                             choices  = cats,
                             selected = cats)   # all selected by default
})

# ── Density plot: before vs after ─────────────────────────────────────────────
output$imp_density_plot <- renderPlotly({
    req(before_after_data())
    df <- before_after_data()
    cols <- c("Before" = "#6c757d", "After" = "#008C94")
    n_na <- sum(is.na(as.numeric(assay(rv$se_filt))))
    p <- plot_ly()
    for (ds in c("Before", "After")) {
        x_vals <- df$expression[df$dataset == ds]
        dens   <- stats::density(x_vals, na.rm = TRUE)
        p <- p %>% add_lines(
            x = dens$x, y = dens$y, name = ds,
            line = list(color = cols[ds], width = 2.5),
            hovertemplate = paste0("<b>", ds, "</b><br>Log\u2082: %{x:.2f}<br>Density: %{y:.4f}<extra></extra>"))
    }
    p %>% layout(
        title  = list(text = paste0("<b>Intensity Density</b>  <span style='font-size:11px;color:#6c757d;'>",
                                    nrow(rv$se_filt), " proteins \u2014 ",
                                    sum(is.na(as.numeric(assay(rv$se_filt)))),
                                    " missing values imputed</span>"),
                      font = list(family="Segoe UI", size=13, color="#002855"), x=0.01),
        xaxis  = list(title="Log\u2082 Intensity", gridcolor="#e9ecef"),
        yaxis  = list(title="Density", gridcolor="#e9ecef"),
        legend = list(orientation="h", y=-0.22, font=list(size=11)),
        paper_bgcolor="white", plot_bgcolor="white", margin=list(t=70,b=65)
    ) %>% config(displayModeBar=TRUE, modeBarButtonsToRemove=c("lasso2d","select2d"),
                 toImageButtonOptions=list(format="png", filename="imputation_density"))
})

# ── Histogram: before vs after ─────────────────────────────────────────────────
output$imp_histogram_plot <- renderPlotly({
    req(before_after_data())
    df <- before_after_data()
    plot_ly(df, x=~expression, color=~dataset,
            colors = c("Before"="#6c757d","After"="#008C94"),
            type="histogram", opacity=0.55, nbinsx=60,
            hovertemplate="<b>%{data.name}</b><br>Intensity: %{x:.2f}<br>Count: %{y}<extra></extra>") %>%
    layout(
        barmode="overlay",
        title  = list(text="<b>Intensity Histogram</b><br><span style='font-size:11px;color:#6c757d;'>Before vs After Imputation</span>",
                      font=list(family="Segoe UI",size=13,color="#002855"), x=0.01),
        xaxis  = list(title="Log\u2082 Intensity", gridcolor="#e9ecef"),
        yaxis  = list(title="Count", gridcolor="#e9ecef"),
        legend = list(orientation="h", y=-0.22),
        paper_bgcolor="white", plot_bgcolor="white", margin=list(t=70,b=65)
    ) %>% config(displayModeBar=TRUE, modeBarButtonsToRemove=c("lasso2d","select2d"),
                 toImageButtonOptions=list(format="png", filename="imputation_histogram"))
})

# ── FC density by imputation category ─────────────────────────────────────────
output$imp_fc_density_plot <- renderPlotly({
    req(imp_fc_data())
    df       <- imp_fc_data()
    sel_cats <- input$imp_cats_filter
    show_sig <- isTRUE(input$imp_show_sig)
    alpha    <- input$alpha; lfc_val <- input$lfc

    # Robust filter: fallback to all categories if sel_cats causes zero overlap
    if (!is.null(sel_cats) && length(sel_cats) > 0) {
        filtered <- df[df$imp_status %in% sel_cats, ]
        if (nrow(filtered) > 0) df <- filtered
    }
    if (show_sig)
        df <- df[!is.na(df$p_adj) & df$p_adj <= alpha & abs(df$log2FC) >= lfc_val, ]
    if (nrow(df) == 0)
        return(plotly_empty() %>%
                   layout(title="No significant proteins at current thresholds \u2014 uncheck filter."))

    cats <- sort(unique(df$imp_status))
    pal  <- setNames(condition_palette(length(cats)), cats)

    # Pre-compute ALL densities into ONE data frame (avoids for-loop plotly build bugs)
    dens_list <- lapply(cats, function(cat) {
        x_vals <- df$log2FC[df$imp_status == cat & !is.na(df$log2FC)]
        if (length(x_vals) < 3) return(NULL)
        d <- tryCatch(stats::density(x_vals, bw = "nrd", adjust = 1),
                      error = function(e) NULL)
        if (is.null(d)) return(NULL)
        n_up <- sum(!is.na(df$p_adj[df$imp_status==cat]) &
                        df$p_adj[df$imp_status==cat]  <= alpha &
                        df$log2FC[df$imp_status==cat] >= lfc_val, na.rm=TRUE)
        n_dn <- sum(!is.na(df$p_adj[df$imp_status==cat]) &
                        df$p_adj[df$imp_status==cat]  <= alpha &
                        df$log2FC[df$imp_status==cat] <= -lfc_val, na.rm=TRUE)
        data.frame(x=d$x, y=d$y,
                   category = paste0(cat," (n=",length(x_vals),", \u2191",n_up," \u2193",n_dn,")"),
                   cat_key  = cat,
                   stringsAsFactors=FALSE)
    })
    dens_df <- do.call(rbind, dens_list[!vapply(dens_list, is.null, logical(1))])

    if (is.null(dens_df) || nrow(dens_df) == 0)
        return(plotly_empty() %>% layout(title="Density could not be computed for current selection."))

    # Map palette to extended names
    cat_pal <- setNames(pal[dens_df$cat_key], dens_df$category)
    cat_pal <- cat_pal[!duplicated(names(cat_pal))]

    subtitle <- if (show_sig) " \u2014 significant only" else " \u2014 all proteins"
    y_max    <- max(dens_df$y) * 1.08

    plot_ly(dens_df, x=~x, y=~y,
            color       = ~category,
            colors      = cat_pal,
            type        = "scatter", mode = "lines",
            line        = list(width=2.3),
            hovertemplate = "<b>%{data.name}</b><br>log\u2082FC: %{x:.2f}<br>Density: %{y:.4f}<extra></extra>"
    ) %>%
    # Vertical reference line at x=0 via add_segments (universally compatible)
    add_segments(x=0, xend=0, y=0, yend=y_max,
                 line=list(color="#ced4da", dash="dot", width=1.3),
                 showlegend=FALSE, hoverinfo="none", inherit=FALSE) %>%
    layout(
        title  = list(text=paste0("<b>log\u2082FC by Imputation Category</b>",
                                  "<br><span style='font-size:11px;color:#6c757d;'>",
                                  input$imp_contrast, subtitle, "</span>"),
                      font=list(family="Segoe UI",size=13,color="#002855"), x=0.01),
        xaxis  = list(title="log\u2082 Fold Change", gridcolor="#e9ecef",
                      zeroline=FALSE),
        yaxis  = list(title="Density", gridcolor="#e9ecef", range=c(0, y_max)),
        legend = list(title=list(text="<b>Category (n, \u2191\u2193 sig)</b>"),
                      orientation="h", y=-0.3, font=list(size=9.5)),
        paper_bgcolor="white", plot_bgcolor="white", margin=list(t=70,b=100)
    ) %>%
    config(displayModeBar=TRUE, modeBarButtonsToRemove=c("lasso2d","select2d"),
           toImageButtonOptions=list(format="png",
                                     filename=paste0("FC_density_",input$imp_contrast)))
})

# ── FC boxplot by imputation category ─────────────────────────────────────────
output$imp_fc_boxplot <- renderPlotly({
    req(imp_fc_data())
    df       <- imp_fc_data()
    sel_cats <- input$imp_cats_filter
    show_sig <- isTRUE(input$imp_show_sig)

    # Same robust filter as density plot
    if (!is.null(sel_cats) && length(sel_cats) > 0) {
        filtered <- df[df$imp_status %in% sel_cats, ]
        if (nrow(filtered) > 0) df <- filtered
    }
    if (show_sig) df <- df[!is.na(df$p_adj) & df$p_adj <= input$alpha & abs(df$log2FC) >= input$lfc, ]
    if (nrow(df) == 0)
        return(plotly_empty() %>%
                   layout(title="No significant proteins at the current thresholds \u2014 uncheck 'Significant proteins only'."))
    cats <- sort(unique(df$imp_status))
    pal  <- setNames(condition_palette(length(cats)), cats)
    df$imp_status <- factor(df$imp_status, levels=cats)
    df$hover <- paste0("<b>", df$gene, "</b><br>Category: ", df$imp_status,
                       "<br>log\u2082FC: ", round(df$log2FC, 3),
                       if ("p_adj" %in% colnames(df)) paste0("<br>adj.p: ", format(df$p_adj, digits=3, scientific=TRUE)) else "")
    subtitle <- if (show_sig) " \u2014 significant only" else " \u2014 all proteins"
    plot_ly(df, x=~imp_status, y=~log2FC, color=~imp_status, colors=pal,
            type="box", text=~hover, hoverinfo="text",
            boxpoints="outliers", marker=list(size=4, opacity=0.55),
            line=list(width=1.5)) %>%
    layout(
        title  = list(text=paste0("<b>log\u2082FC Boxplot by Imputation Category</b>",
                                  "<br><span style='font-size:11px;color:#6c757d;'>",
                                  input$imp_contrast, subtitle, "</span>"),
                      font=list(family="Segoe UI",size=13,color="#002855"), x=0.01),
        xaxis  = list(title="Imputation (numerator\u2212denominator)",
                      gridcolor="#e9ecef", tickangle=-35),
        yaxis  = list(title="log\u2082 Fold Change", gridcolor="#e9ecef",
                      zeroline=TRUE, zerolinecolor="#ced4da"),
        showlegend=FALSE,
        paper_bgcolor="white", plot_bgcolor="white", margin=list(t=70,b=110)
    ) %>% config(displayModeBar=TRUE, modeBarButtonsToRemove=c("lasso2d","select2d"),
                 toImageButtonOptions=list(format="png", filename=paste0("FC_boxplot_",input$imp_contrast)))
})

# ── Imputation type summary DT ─────────────────────────────────────────────────
output$imp_type_table <- DT::renderDT({
    req(imp_summary_data())
    tbl <- imp_summary_data()
    colnames(tbl) <- c("Category (num\u2212denom)", "N proteins",
                       "\u2191 Up", "\u2193 Down", "Median log\u2082FC")
    DT::datatable(tbl, rownames=FALSE,
        class    = "table table-sm table-hover",
        extensions = "Buttons",
        options  = list(
            dom="Bt", ordering=TRUE,
            buttons = list(
                list(extend="csv",   text="CSV",
                     filename=paste0("imp_summary_",Sys.Date())),
                list(extend="excel", text="Excel",
                     filename=paste0("imp_summary_",Sys.Date()))
            ),
            pageLength=20
        )
    ) %>%
    DT::formatStyle("Category (num\u2212denom)", fontWeight="bold", color="#002855") %>%
    DT::formatStyle("\u2191 Up",
        color="white",
        backgroundColor=DT::styleInterval(0, c("transparent","#dc3545")),
        fontWeight="bold") %>%
    DT::formatStyle("\u2193 Down",
        color="white",
        backgroundColor=DT::styleInterval(0, c("transparent","#0d6efd")),
        fontWeight="bold") %>%
    DT::formatStyle("Median log\u2082FC",
        color=DT::styleInterval(c(-0.5, 0.5), c("#0d6efd","#333","#dc3545")),
        fontWeight="bold")
}, server=FALSE)

# ── Filtered proteins DT ───────────────────────────────────────────────────────
output$imp_filtered_table <- DT::renderDT({
    req(rv$removed)
    tbl <- rv$removed
    colnames(tbl) <- c("Protein ID", "Gene", "Reason")
    DT::datatable(tbl, rownames=FALSE,
        class      = "table table-sm table-hover",
        filter     = "top",
        extensions = "Buttons",
        options    = list(
            dom        = "Bfrtip",
            pageLength = 15,
            scrollX    = TRUE,
            buttons    = list(
                list(extend="csv",  text="CSV",
                     filename=paste0("filtered_proteins_",Sys.Date())),
                list(extend="excel",text="Excel",
                     filename=paste0("filtered_proteins_",Sys.Date()))
            )
        )
    ) %>%
    DT::formatStyle("Protein ID", fontWeight="bold", color="#002855")
}, server=TRUE)

# ── Download handlers: raw matrices ───────────────────────────────────────────
output$dl_before_matrix <- downloadHandler(
    filename = function() paste0(input$experiment, "_before_imputation_", Sys.Date(), ".csv"),
    content  = function(f) {
        req(rv$se_filt)
        rd  <- as.data.frame(rowData(rv$se_filt))
        mat <- as.data.frame(assay(rv$se_filt))
        g   <- if ("Genes" %in% colnames(rd)) rd$Genes else rownames(rd)
        write.csv(cbind(Gene=g, ProteinGroup=rownames(rd), mat), f, row.names=FALSE)
    }
)
output$dl_after_matrix <- downloadHandler(
    filename = function() paste0(input$experiment, "_after_imputation_", Sys.Date(), ".csv"),
    content  = function(f) {
        req(rv$se_imp)
        rd  <- as.data.frame(rowData(rv$se_imp))
        mat <- as.data.frame(assay(rv$se_imp))
        g   <- if ("Genes" %in% colnames(rd)) rd$Genes else rownames(rd)
        write.csv(cbind(Gene=g, ProteinGroup=rownames(rd), mat), f, row.names=FALSE)
    }
)

# ════════════════════════════════════════════════════════════════════════════════
# DIFFERENTIAL EXPRESSION — server logic
# ════════════════════════════════════════════════════════════════════════════════

observe({
    req(rv$status == "done", rv$tests)
    updateSelectInput(session, "de_contrast",
                      choices = rv$tests, selected = rv$tests[1])
})

observe({
    req(rv$status == "done")
    has_int <- !identical(rv$tests_interaction, "NA")
    ch <- c("All proteins" = "all_common_effect",
            "Common (stable) effect" = "common_effect")
    if (has_int) ch["Interaction effect"] <- "interaction_effect"
    updateSelectInput(session, "de_effect_type", choices = ch, selected = "all_common_effect")
})

# ── Determine contrast column for a given effect slot ─────────────────────────
.de_contrast_col <- function(ct, effect_slot, tests, tests_interaction) {
    if (effect_slot == "interaction_effect" && !identical(tests_interaction, "NA")) {
        idx <- which(tests == ct)
        if (length(idx) > 0 && idx <= length(tests_interaction))
            return(tests_interaction[idx])
    }
    ct
}

# ── Volcano data ───────────────────────────────────────────────────────────────
de_volcano_data <- reactive({
    req(rv$status == "done", input$de_contrast, rv$effects)
    ct   <- input$de_contrast
    slot <- if (!is.null(input$de_effect_type)) input$de_effect_type else "all_common_effect"
    se_obj <- rv$effects[[ct]][[slot]]
    if (is.null(se_obj) || nrow(se_obj) == 0) return(NULL)
    rd <- as.data.frame(rowData(se_obj))

    plot_ct  <- .de_contrast_col(ct, slot, rv$tests, rv$tests_interaction)
    lfc_col  <- paste0("log2FC_", plot_ct)
    padj_col <- paste0("p.adj_",  plot_ct)

    # Fallback: search for any matching log2FC column
    if (!lfc_col %in% colnames(rd)) {
        cands <- grep("^log2FC_", colnames(rd), value=TRUE)
        if (!length(cands)) return(NULL)
        lfc_col  <- cands[1]
        padj_col <- sub("^log2FC_", "p.adj_", lfc_col)
    }
    if (!padj_col %in% colnames(rd)) {
        padj_col <- sub("^log2FC_", "p.val_", lfc_col)
        if (!padj_col %in% colnames(rd)) return(NULL)
    }

    gene_col <- Filter(function(x) x %in% colnames(rd), c("Genes","name","ID"))[1]
    prot_col <- Filter(function(x) x %in% colnames(rd), c("Protein.Names","Protein.Group","ID"))[1]

    df <- data.frame(
        protein_id = rownames(rd),
        gene       = if (!is.na(gene_col)) rd[[gene_col]] else rownames(rd),
        prot_name  = if (!is.na(prot_col)) rd[[prot_col]] else rownames(rd),
        log2FC     = rd[[lfc_col]],
        p_adj      = rd[[padj_col]],
        imputed    = if ("imputed" %in% colnames(rd)) as.logical(rd$imputed) else FALSE,
        stringsAsFactors = FALSE
    )
    df <- df[!is.na(df$log2FC), , drop=FALSE]
    df$p_adj[is.na(df$p_adj)] <- NA_real_
    df$neg_log10_p <- -log10(pmax(df$p_adj, 1e-300))

    alpha <- input$alpha; lfc_v <- input$lfc
    df$sig_class <- ifelse(
        !is.na(df$p_adj) & df$p_adj <= alpha & abs(df$log2FC) >= lfc_v,
        ifelse(df$imputed, "Significant (imputed)", "Significant"),
        "Not significant"
    )
    attr(df, "plot_ct") <- plot_ct
    df
})

# ── Volcano plot ───────────────────────────────────────────────────────────────
output$de_volcano_plot <- renderPlotly({
    req(de_volcano_data())
    df      <- de_volcano_data()
    ct      <- input$de_contrast
    plot_ct <- attr(df, "plot_ct")
    alpha   <- input$alpha; lfc_v <- input$lfc
    n_lbl   <- if (!is.null(input$de_n_labels)) as.integer(input$de_n_labels) else 20L

    color_map <- c("Significant"="#dc3545",
                   "Significant (imputed)"="#fd7e14",
                   "Not significant"="#ced4da")

    df$hover <- paste0(
        "<b>", df$gene, "</b><br>",
        df$prot_name, "<br>",
        "log\u2082FC: <b>", round(df$log2FC,3), "</b><br>",
        "adj.p: <b>", format(df$p_adj, digits=3, scientific=TRUE), "</b>",
        ifelse(df$imputed,
               "<br><span style='color:#fd7e14;'>\u26a0 imputed</span>", "")
    )

    x_rng  <- max(abs(df$log2FC), na.rm=TRUE) * 1.06
    y_max  <- max(df$neg_log10_p[is.finite(df$neg_log10_p)], na.rm=TRUE) * 1.06
    thr_y  <- -log10(alpha)

    p <- plot_ly()
    for (cls in c("Not significant", "Significant (imputed)", "Significant")) {
        sub <- df[df$sig_class == cls, ]
        if (!nrow(sub)) next
        p <- p %>% add_trace(
            data=sub, x=~log2FC, y=~neg_log10_p, text=~hover,
            type="scatter", mode="markers", name=cls,
            hoverinfo="text",
            marker=list(color=color_map[cls],
                        size=ifelse(cls=="Not significant",5,7),
                        opacity=ifelse(cls=="Not significant",0.4,0.85),
                        line=if(cls=="Not significant") list(width=0) else
                                 list(color="white",width=0.5))
        )
    }

    # Threshold lines via add_segments (universally compatible)
    p <- p %>%
        add_segments(x=-x_rng, xend=x_rng, y=thr_y, yend=thr_y,
                     line=list(color="#6c757d",dash="dot",width=1.1),
                     showlegend=FALSE, hoverinfo="none", inherit=FALSE) %>%
        add_segments(x=lfc_v,  xend=lfc_v,  y=0, yend=y_max,
                     line=list(color="#6c757d",dash="dot",width=1.1),
                     showlegend=FALSE, hoverinfo="none", inherit=FALSE) %>%
        add_segments(x=-lfc_v, xend=-lfc_v, y=0, yend=y_max,
                     line=list(color="#6c757d",dash="dot",width=1.1),
                     showlegend=FALSE, hoverinfo="none", inherit=FALSE)

    # Gene labels for top N significant
    sig_df <- df[df$sig_class != "Not significant" & !is.na(df$p_adj), ]
    sig_df <- sig_df[order(sig_df$p_adj), ]
    if (n_lbl > 0 && nrow(sig_df) > 0) {
        lbl_df <- sig_df[seq_len(min(n_lbl, nrow(sig_df))), ]
        p <- p %>% add_annotations(
            data=lbl_df, x=~log2FC, y=~neg_log10_p, text=~gene,
            showarrow=TRUE, arrowhead=2, arrowsize=0.7, arrowcolor="#555",
            font=list(size=9.5, color="#002855"),
            bgcolor="rgba(255,255,255,0.75)", borderpad=2
        )
    }

    n_up  <- sum(df$sig_class != "Not significant" & df$log2FC > 0, na.rm=TRUE)
    n_dn  <- sum(df$sig_class != "Not significant" & df$log2FC < 0, na.rm=TRUE)
    effect_lbl <- switch(input$de_effect_type,
                         all_common_effect  = "All proteins",
                         common_effect      = "Common effect",
                         interaction_effect = "Interaction effect",
                         "All proteins")

    p %>% layout(
        title = list(
            text=paste0("<b>Volcano: ", ct, "</b>  ",
                        "<span style='font-size:11px;color:#6c757d;'>",
                        effect_lbl, " \u2014 \u2191",n_up," / \u2193",n_dn," significant</span>"),
            font=list(family="Segoe UI",size=14,color="#002855"), x=0.01),
        xaxis = list(title="log\u2082 Fold Change", gridcolor="#e9ecef",
                     zeroline=TRUE, zerolinecolor="#dee2e6"),
        yaxis = list(title="-log\u2081\u2080(adj.p)", gridcolor="#e9ecef"),
        legend= list(orientation="h", y=-0.18, font=list(size=11)),
        paper_bgcolor="white", plot_bgcolor="white",
        margin=list(t=75, b=75)
    ) %>%
        config(displayModeBar=TRUE,
               modeBarButtonsToRemove=c("lasso2d","select2d","autoScale2d"),
               toImageButtonOptions=list(format="png",
                                         filename=paste0("volcano_",ct)))
})

# ── Deregulogram data ──────────────────────────────────────────────────────────
de_dereg_data <- reactive({
    req(rv$status == "done", input$de_contrast, rv$se_limma)
    if (identical(rv$tests_interaction, "NA")) return(NULL)
    ct    <- input$de_contrast
    idx   <- which(rv$tests == ct)
    if (!length(idx)) return(NULL)
    inter_ct <- rv$tests_interaction[min(idx, length(rv$tests_interaction))]
    if (identical(inter_ct, "NA")) return(NULL)
    rd <- as.data.frame(rowData(rv$se_limma))
    cols_need <- c(paste0("log2FC_",ct), paste0("p.adj_",ct),
                   paste0("log2FC_",inter_ct), paste0("p.adj_",inter_ct))
    if (!all(cols_need %in% colnames(rd))) return(NULL)
    gene_col <- Filter(function(x) x %in% colnames(rd), c("Genes","name","ID"))[1]
    df <- data.frame(
        gene   = if (!is.na(gene_col)) rd[[gene_col]] else rownames(rd),
        fc_ref = rd[[paste0("log2FC_",ct)]],
        fc_int = rd[[paste0("log2FC_",inter_ct)]],
        p_ref  = rd[[paste0("p.adj_",ct)]],
        p_int  = rd[[paste0("p.adj_",inter_ct)]],
        stringsAsFactors=FALSE
    )
    df <- df[complete.cases(df[,c("fc_ref","fc_int","p_ref","p_int")]), , drop=FALSE]
    if (!nrow(df)) return(NULL)
    df$fc_lvl2 <- df$fc_ref + df$fc_int
    a  <- input$alpha; l <- input$lfc
    df$sig_class <- ifelse(
        !is.na(df$p_ref) & df$p_ref<=a & !is.na(df$p_int) & df$p_int<=a &
            abs(df$fc_ref)>=l & abs(df$fc_lvl2)>=l & abs(df$fc_int)>=l,
        "sig_interaction",
        ifelse(!is.na(df$p_ref) & df$p_ref<=a &
                   (abs(df$fc_ref)>=l | abs(df$fc_lvl2)>=l),
               "sig_no_interaction", "non_sig"))
    list(df=df, main_ct=ct, inter_ct=inter_ct)
})

# ── Deregulogram plot ──────────────────────────────────────────────────────────
output$de_dereg_plot <- renderPlotly({
    req(de_dereg_data())
    dat   <- de_dereg_data()
    df    <- dat$df
    info  <- tryCatch(
        ProteinBatcher:::.pp_infer_interaction_factor(rv$se_limma, rv$tests_interaction),
        error=function(e) NULL)
    fac_name <- if (!is.null(info)) info$name   else "factor"
    fac_lev  <- if (!is.null(info)) info$levels else c("ref","other")
    col_map  <- c(sig_interaction="dc3545",sig_no_interaction="0d6efd",non_sig="ced4da")
    hex_col  <- paste0("#", col_map)
    names(hex_col) <- names(col_map)
    lbl_map  <- c(sig_interaction   = paste0("Sig. \u00d7 ", fac_name),
                  sig_no_interaction= "Sig. condition only",
                  non_sig           = "Not significant")
    df$hover <- paste0(
        "<b>",df$gene,"</b>",
        "<br>",fac_lev[1],": ",round(df$fc_ref,3),
        "<br>",fac_lev[2],": ",round(df$fc_lvl2,3),
        "<br>Interaction: ",round(df$fc_int,3),
        "<br>",lbl_map[df$sig_class])

    xy_rng <- max(abs(c(df$fc_ref, df$fc_lvl2)),na.rm=TRUE)*1.08
    p <- plot_ly()
    for (cls in c("non_sig","sig_no_interaction","sig_interaction")) {
        s <- df[df$sig_class==cls, ]
        if (!nrow(s)) next
        p <- p %>% add_trace(
            data=s, x=~fc_lvl2, y=~fc_ref, text=~hover,
            type="scatter", mode="markers", name=lbl_map[cls], hoverinfo="text",
            marker=list(color=hex_col[cls],
                        size=ifelse(cls=="non_sig",5,7),
                        opacity=ifelse(cls=="non_sig",0.32,0.85),
                        line=if(cls=="non_sig") list(width=0) else
                                 list(color="white",width=0.4)))
    }
    # Identity diagonal + cross-hairs
    p <- p %>%
        add_segments(x=-xy_rng,xend=xy_rng,y=-xy_rng,yend=xy_rng,
                     line=list(color="#adb5bd",dash="dash",width=1),
                     showlegend=FALSE,hoverinfo="none",inherit=FALSE) %>%
        add_segments(x=-xy_rng,xend=xy_rng,y=0,yend=0,
                     line=list(color="#dee2e6",dash="dot",width=0.7),
                     showlegend=FALSE,hoverinfo="none",inherit=FALSE) %>%
        add_segments(x=0,xend=0,y=-xy_rng,yend=xy_rng,
                     line=list(color="#dee2e6",dash="dot",width=0.7),
                     showlegend=FALSE,hoverinfo="none",inherit=FALSE)
    # Label top significant interaction proteins
    sig_int <- df[df$sig_class=="sig_interaction", ]
    n_ann   <- min(20L, nrow(sig_int))
    if (n_ann > 0) {
        top_ann <- sig_int[order(abs(sig_int$fc_int), decreasing=TRUE)[seq_len(n_ann)], ]
        p <- p %>% add_annotations(
            data=top_ann, x=~fc_lvl2, y=~fc_ref, text=~gene,
            showarrow=TRUE, arrowhead=2, arrowsize=0.6, arrowcolor="#555",
            font=list(size=9,color="#002855"),
            bgcolor="rgba(255,255,255,0.8)", borderpad=2)
    }
    p %>% layout(
        title=list(text=paste0("<b>Deregulogram: ",dat$main_ct,"</b>",
                               "<br><span style='font-size:11px;color:#6c757d;'>",
                               fac_name,": ",fac_lev[1]," (y) vs ",fac_lev[2]," (x)</span>"),
                   font=list(family="Segoe UI",size=13,color="#002855"),x=0.01),
        xaxis=list(title=paste0("Full effect (",fac_lev[2],")"),gridcolor="#e9ecef",
                   zeroline=TRUE,zerolinecolor="#dee2e6"),
        yaxis=list(title=paste0("Full effect (",fac_lev[1],")"),gridcolor="#e9ecef",
                   zeroline=TRUE,zerolinecolor="#dee2e6"),
        legend=list(orientation="h",y=-0.2,font=list(size=10)),
        paper_bgcolor="white",plot_bgcolor="white",margin=list(t=70,b=70)
    ) %>% config(displayModeBar=TRUE,
                 modeBarButtonsToRemove=c("lasso2d","select2d","autoScale2d"),
                 toImageButtonOptions=list(format="png",
                                           filename=paste0("deregulogram_",dat$main_ct)))
})

# ── Conditional deregulogram panel ────────────────────────────────────────────
output$de_dereg_panel_ui <- renderUI({
    req(rv$status == "done")
    if (identical(rv$tests_interaction, "NA"))
        return(tags$div(class="text-center text-muted p-4",
                        icon("circle-info"),
                        " Deregulogram not available: no interaction contrasts were tested.",
                        tags$br(),
                        tags$small("Include interaction tests in your contrasts file and enable them in Parameters.")))
    dat <- de_dereg_data()
    if (is.null(dat))
        return(tags$div(class="text-center text-muted p-3",
                        icon("triangle-exclamation"),
                        " Required columns not found in rowData. Ensure the selected contrast has interaction results."))
    info  <- tryCatch(
        ProteinBatcher:::.pp_infer_interaction_factor(rv$se_limma, rv$tests_interaction),
        error=function(e) NULL)
    fac_name <- if (!is.null(info)) info$name else "interaction factor"
    fac_lev  <- if (!is.null(info)) paste(info$levels, collapse=" / ") else "level1 / level2"

    tagList(
        tags$p(style="font-size:.83em;color:#6c757d;margin-bottom:8px;",
               icon("circle-info"),
               paste0(" Factor: ", tags$b(fac_name), " (",fac_lev,")."),
               " Red = significant interaction; blue = condition-only; grey = not significant."),
        plotlyOutput("de_dereg_plot", height="420px"),
        tags$p(style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;",
               tags$b("Interpretation:"), " Proteins along the diagonal behave consistently across",
               " both factor levels. Off-diagonal proteins show factor-dependent regulation.",
               " The deregulogram is more restrictive than the interaction volcano — it requires",
               " a significant interaction AND a biologically relevant full effect in at least one level.")
    )
})

# ── DEG table ──────────────────────────────────────────────────────────────────
output$de_deg_table <- DT::renderDT({
    req(de_volcano_data())
    df       <- de_volcano_data()
    alpha    <- input$alpha; lfc_v <- input$lfc
    sig_df   <- df[df$sig_class != "Not significant", , drop=FALSE]
    gf       <- trimws(if (!is.null(input$de_gene_filter)) input$de_gene_filter else "")
    if (nchar(gf) > 0)
        sig_df <- sig_df[grepl(gf, sig_df$gene, ignore.case=TRUE) |
                         grepl(gf, sig_df$protein_id, ignore.case=TRUE), , drop=FALSE]
    if (!nrow(sig_df)) {
        return(DT::datatable(
            data.frame(Note="No significant proteins at current thresholds / filter."),
            rownames=FALSE, options=list(dom="t")))
    }
    sig_df  <- sig_df[order(sig_df$p_adj, na.last=TRUE), ]
    ct_lbl  <- input$de_contrast
    tbl <- data.frame(
        Gene      = sig_df$gene,
        ProteinID = sig_df$protein_id,
        log2FC    = round(sig_df$log2FC, 4),
        adj.p     = format(sig_df$p_adj, digits=3, scientific=TRUE),
        "-log10p" = round(sig_df$neg_log10_p, 2),
        Direction = ifelse(sig_df$log2FC > 0, "\u2191 Up", "\u2193 Down"),
        Imputed   = sig_df$imputed,
        check.names=FALSE, stringsAsFactors=FALSE
    )
    DT::datatable(tbl, rownames=FALSE,
        class="table table-sm table-hover", filter="top",
        extensions="Buttons",
        options=list(
            dom="Bfrtip", pageLength=20, scrollX=TRUE,
            buttons=list(
                list(extend="csv",   text="CSV",
                     filename=paste0("DEGs_",ct_lbl,"_",Sys.Date())),
                list(extend="excel", text="Excel",
                     filename=paste0("DEGs_",ct_lbl,"_",Sys.Date()))
            )
        )
    ) %>%
        DT::formatStyle("Direction",
            color=DT::styleEqual(c("\u2191 Up","\u2193 Down"),c("#dc3545","#0d6efd")),
            fontWeight="bold") %>%
        DT::formatStyle("log2FC",
            color=DT::styleInterval(c(-0.001,0.001),c("#0d6efd","#333","#dc3545")),
            fontWeight="bold") %>%
        DT::formatStyle("Gene", fontWeight="bold", color="#002855") %>%
        DT::formatStyle("Imputed",
            color=DT::styleEqual(c("TRUE","FALSE"),c("#fd7e14","#6c757d")))
}, server=TRUE)
```

## Overview

::: {.layout-ncol=2}

::: {.card}
### ProteinBatcher Visualizer {.card-title}

Standardized, design-aware downstream analysis of **label-free quantitative proteomics** (DIA-NN outputs). Upload your data, define contrasts, and run the full pipeline in a single click.

---

**Pipeline steps**

1. **Import** — reads your protein matrix and annotation into a `SummarizedExperiment`.
2. **Filter** — removes proteins below the missingness threshold in all conditions.
3. **Impute** — two-step, condition-aware strategy: *mean imputation* for partially missing proteins; *LDV (left-censored draw)* for proteins missing across an entire condition.
4. **Differential testing** — `limma` with imputation-aware BH-FDR: LDV/LDV features are excluded from p-value adjustment.
5. **Effect organisation** — splits results into `all_common_effect`, `common_effect` and `interaction_effect` per contrast.

---

**Getting started**

1. Upload annotation and DIA-NN protein matrix.
2. Upload contrasts — use the example file as a template.
3. Adjust parameters, then click **Run ProteinBatcher**.
4. Explore results in the tabs above.

---

**Contrasts file format**

```
[design]
formula = ~ 0 + condition + batch + condition:batch
reference_condition = NoTreated

[tests]
IL13_vs_NoTreated
IL22_vs_NoTreated

[interaction_tests]
IL13.batchday2
IL22.batchday2
```
:::

::: {.card}
### Pipeline status {.card-title}

```{r}
uiOutput("pipeline_summary_ui")
```
:::

:::

::: {.card}
### Example outputs {.card-title}

::: {.columns}
::: {.column width="50%"}
**Main Effect — all proteins**
```{r}
#| echo: false
htmltools::img(
    src   = "proteinbatcher-extdata/HaCaT_MainEffect_AllProteins_IL13_vs_NoTreated.png",
    style = "max-width:100%;border-radius:6px;border:1px solid #dee2e6;"
)
```
Volcano plot showing the global IL13 effect across all tested proteins.
:::
::: {.column width="50%"}
**Deregulogram — batch-level comparison**
```{r}
#| echo: false
htmltools::img(
    src   = "proteinbatcher-extdata/HaCat_Deregulogram_IL13_vs_NoTreated_batch.png",
    style = "max-width:100%;border-radius:6px;border:1px solid #dee2e6;"
)
```
Full IL13 effects in batch day 1 (y) vs day 2 (x). Proteins off the diagonal show batch-dependent responses.
:::
:::
:::

::: {.card}
### Interpretation & Support {.card-title}

::: {.layout-ncol=3}

::: {}
**Interpreting results**

- **All proteins volcano** — global condition effect; includes batch-dependent proteins.
- **Common effect** — proteins with a consistent condition response. Most biologically robust.
- **Interaction effect** — proteins whose response differs significantly between factor levels.
- **Deregulogram** — emphasises direction-changing or magnitude-shifting proteins. More restrictive than the interaction volcano.
:::

::: {}
**Support**

- **Bug reports / features:** [GitHub issues](https://github.com/DataScienceRD-Almirall/ProteinBatcher).
- **Documentation:** see the ProteinBatcher vignette.
- **Contrasts file:** download the example as a template. Comments (`#`) and blank lines are ignored.
- **File size limit:** max 100 MB per file.
:::

::: {}
**About**

Developed by **Aitor Moruno-Cuenca** with contributions from the Almirall Data Science R&D team (Dr. Sergi Sayols-Puig, Dr. Bruna Oriol-Tordera, Dr. Francesc Fernández-Albert, Dr. Elisa Monzón Casanova) and the CRG/UPF Proteomics Unit (Dr. Hiba Salim, Dr. Guadalupe Espadas-García, Prof. Dr. Eduard Sabidó-Aguade).

Supervised by Prof. Dr. Alexandre Perera-Lluna (UPC).

*ProteinBatcher* v0.99.0 — MIT License.
:::

:::
:::

# Data inspection

## {.sidebar}

```{r di-sidebar}
tags$div(
    style = "padding:4px 0;",

    # PCA controls
    tags$div(
        style = "font-size:.73em;font-weight:700;text-transform:uppercase;
                 letter-spacing:.08em;color:#00F0BE;margin-bottom:8px;",
        icon("circle-nodes"), " PCA options"),
    selectInput("di_color_by", "Colour samples by",
                choices = c("condition"), selected = "condition"),
    selectInput("di_shape_by", "Shape samples by",
                choices = c("None"), selected = "None"),
    tags$div(style = "display:flex;gap:6px;",
        tags$div(style="flex:1;",
            selectInput("di_pc_x", "X axis",
                        choices = paste0("PC", 1:5), selected = "PC1")),
        tags$div(style="flex:1;",
            selectInput("di_pc_y", "Y axis",
                        choices = paste0("PC", 1:5), selected = "PC2"))
    ),

    tags$hr(style = "border-color:#ffffff30;margin:14px 0;"),

    # Protein explorer controls
    tags$div(
        style = "font-size:.73em;font-weight:700;text-transform:uppercase;
                 letter-spacing:.08em;color:#00F0BE;margin-bottom:8px;",
        icon("magnifying-glass"), " Protein Explorer"),
    selectizeInput(
        "di_gene", "Search gene / protein",
        choices = NULL,
        options = list(
            placeholder    = "Type gene name (e.g. SERPINB3)\u2026",
            onInitialize   = I('function() { this.setValue(""); }'),
            maxOptions     = 200,
            searchField    = c("label", "value"),
            highlight      = TRUE
        )
    ),
    selectInput("di_dot_color", "Colour dots by",
                choices = c("condition"), selected = "condition"),
    selectInput("di_dot_shape", "Shape dots by",
                choices = c("None"), selected = "None"),

    tags$hr(style = "border-color:#ffffff30;margin:14px 0;"),

    # Summary table controls
    tags$div(
        style = "font-size:.73em;font-weight:700;text-transform:uppercase;
                 letter-spacing:.08em;color:#00F0BE;margin-bottom:8px;",
        icon("table"), " Summary Table"),
    selectInput("di_contrast", "Contrast for DE counts",
                choices = NULL, selected = NULL),
    tags$div(class="text-muted mt-1", style="font-size:.78em;line-height:1.4;",
             icon("circle-info"),
             " Uses \u03b1 and log\u2082FC thresholds from Parameters.")
)
```

## Row {height="98px"}

```{r}
uiOutput("di_metrics_ui")
```

## Row {height="470px"}

::: {.card title="Principal Component Analysis"}

```{r}
plotlyOutput("di_pca_plot", height = "370px")
```

<p style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;">
Each point is a sample. Samples clustering together share similar global protein abundance profiles.
Separation by condition indicates a detectable treatment effect; separation by batch or donor points to a technical or biological covariate worth including in the model.
</p>
:::

::: {.card title="Scree — Variance Explained" width="32%"}

```{r}
plotlyOutput("di_scree_plot", height = "350px")
```

<p style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;">
Bars: variance per PC. Dotted line: cumulative total.
A steep early drop-off suggests the biological signal is concentrated in the first few PCs.
</p>
:::

## Row {height="520px"}

::: {.card title="Protein Abundance Explorer" width="60%"}

```{r}
tags$p(style="font-size:.83em;color:#6c757d;margin-bottom:8px;",
       icon("circle-info"),
       " Search for a protein above. Open circles = imputed values.",
       " Thick bar = condition mean.")
plotlyOutput("di_protein_plot", height = "410px")
```

<p style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;">
Individual sample abundances (Log\u2082) are shown per condition; the thick bar marks the condition mean.
Open circles are imputed values (mean or LDV strategy) — interpret with caution for proteins with high missingness.
</p>
:::

::: {.card title="Differential Expression Summary" width="40%"}

```{r}
tags$p(style="font-size:.83em;color:#6c757d;margin-bottom:8px;",
       icon("circle-info"),
       " Up/down counts apply the \u03b1 and log\u2082FC thresholds set in Parameters.")
DT::DTOutput("di_summary_table")
```

<p style="font-size:.82em;color:#6c757d;margin-top:8px;padding:0 2px;">
<b>\u22651 imputed</b>: proteins with at least one imputed value in either group.
LDV/LDV pairs are excluded from BH correction and will have <em>NA</em> adjusted p-values.
</p>
:::

# Imputation and filtering outputs

## {.sidebar}

```{r imp-sidebar}
tags$div(
    style = "padding:4px 0;",

    # ── FC Analysis controls ─────────────────────────────────────────────────
    tags$div(style="font-size:.73em;font-weight:700;text-transform:uppercase;
                    letter-spacing:.08em;color:#00F0BE;margin-bottom:8px;",
             icon("chart-bar"), " FC Analysis"),
    selectInput("imp_contrast", "Select contrast",
                choices=NULL, selected=NULL),
    checkboxInput("imp_show_sig",
                  tagList(icon("filter"), " Significant proteins only"),
                  value=FALSE),
    tags$div(class="text-muted mt-1", style="font-size:.78em;line-height:1.4;",
             icon("circle-info"),
             " Significance: adj.p \u2264 \u03b1 and |log\u2082FC| \u2265 LFC (from Parameters)"),

    tags$hr(style="border-color:#ffffff30;margin:14px 0;"),

    # ── Category filter ──────────────────────────────────────────────────────
    tags$div(style="font-size:.73em;font-weight:700;text-transform:uppercase;
                    letter-spacing:.08em;color:#00F0BE;margin-bottom:6px;",
             icon("tag"), " Imputation Categories"),
    tags$div(class="text-muted", style="font-size:.77em;margin-bottom:6px;",
             "Show in FC plots (numerator\u2212denominator):"),
    checkboxGroupInput("imp_cats_filter", NULL, choices=NULL),

    tags$hr(style="border-color:#ffffff30;margin:14px 0;"),

    # ── Downloads ────────────────────────────────────────────────────────────
    tags$div(style="font-size:.73em;font-weight:700;text-transform:uppercase;
                    letter-spacing:.08em;color:#00F0BE;margin-bottom:8px;",
             icon("download"), " Download Matrices"),
    downloadButton("dl_before_matrix",
        label=tagList(icon("file-csv"), " Before imputation"),
        class="btn btn-sm btn-outline-light w-100",
        style="margin-bottom:6px;font-size:.82em;"),
    downloadButton("dl_after_matrix",
        label=tagList(icon("file-csv"), " After imputation"),
        class="btn btn-sm btn-outline-light w-100",
        style="font-size:.82em;"),
    tags$div(class="text-muted mt-2", style="font-size:.77em;line-height:1.4;",
             icon("circle-info"),
             " Tables in the rows below also have their own CSV/Excel download buttons.")
)
```

## Row {height="420px"}

::: {.card title="Intensity Density — Before vs After Imputation"}

```{r}
plotlyOutput("imp_density_plot", height="330px")
```

<p style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;">
LDV imputed values are random draws from the low-intensity tail — they appear as a <b>left-shifted shoulder</b> in the After distribution.
A large separation or a prominent second mode indicates extensive LDV imputation; mean-imputed values blend into the existing distribution.
</p>
:::

::: {.card title="Intensity Histogram — Before vs After Imputation"}

```{r}
plotlyOutput("imp_histogram_plot", height="330px")
```

<p style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;">
Overlapping bars show the absolute count of intensity values before (grey) and after (teal) imputation.
Bars appearing only in the After distribution at lower intensities correspond exclusively to imputed values.
</p>
:::

## Row {height="470px"}

::: {.card title="log₂FC Distribution by Imputation Category — Density"}

```{r}
tags$p(style="font-size:.83em;color:#6c757d;margin-bottom:8px;",
       icon("circle-info"),
       " Use the Category Filter in the sidebar to focus on specific imputation patterns.",
       " Hover each curve for per-category counts and DE numbers.")
plotlyOutput("imp_fc_density_plot", height="370px")
```

<p style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;">
<b>Interpretation:</b> Categories with LDV in one group and nothing in the other (<em>ldv-none</em>, <em>none-ldv</em>) typically show shifted distributions
— this is expected and biologically meaningful (proteins absent in one condition).
<em>ldv-ldv</em> proteins should show <b>no significant results</b> (excluded from BH correction by design).
A shift in <em>mean-mean</em> or <em>none-none</em> would indicate a genuine condition effect.
</p>
:::

::: {.card title="log₂FC Distribution by Imputation Category — Boxplot" width="45%"}

```{r}
plotlyOutput("imp_fc_boxplot", height="370px")
```

<p style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;">
Box spans IQR, whiskers extend to 1.5×IQR; outliers shown as individual points.
Hover an outlier to identify the protein. Large median shifts in <em>none-none</em> confirm genuine biology.
</p>
:::

## Row

::: {.card title="Imputation Category Summary Table" width="50%"}

```{r}
tags$p(style="font-size:.83em;color:#6c757d;margin-bottom:8px;",
       icon("circle-info"),
       " Download buttons: CSV and Excel. Column headers are sortable.")
DT::DTOutput("imp_type_table")
```

<p style="font-size:.82em;color:#6c757d;margin-top:8px;padding:0 2px;">
<b>Category</b>: imputation type in the numerator condition vs denominator condition.
<b>ldv-ldv</b>: excluded from BH correction → should have 0 significant results (verify here).
High up/down counts in asymmetric categories (<em>ldv-none</em>, <em>none-ldv</em>) reflect proteins
present only in one condition and are expected.
</p>
:::

::: {.card title="Proteins Removed by Missingness Filter" width="50%"}

```{r}
tags$p(style="font-size:.83em;color:#6c757d;margin-bottom:8px;",
       icon("circle-info"),
       " These proteins failed the missingness threshold in all conditions.",
       " Use the column search boxes to filter. Download available above.")
DT::DTOutput("imp_filtered_table")
```

<p style="font-size:.82em;color:#6c757d;margin-top:8px;padding:0 2px;">
A protein is removed if it does not reach the <b>minimum observed proportion</b> (set in Parameters) in
<em>any</em> condition. Increasing the threshold retains more proteins but with higher missingness.
Check this list to ensure no biologically relevant proteins are systematically absent.
</p>
:::

# Differential expression analysis

## {.sidebar}

```{r de-sidebar}
tags$div(
    style = "padding:4px 0;",

    tags$div(style="font-size:.73em;font-weight:700;text-transform:uppercase;
                    letter-spacing:.08em;color:#00F0BE;margin-bottom:8px;",
             icon("chart-line"), " Volcano Plot"),
    selectInput("de_contrast",    "Select contrast",    choices=NULL, selected=NULL),
    selectInput("de_effect_type", "Effect type",
                choices=c("All proteins"="all_common_effect"), selected="all_common_effect"),
    numericInput("de_n_labels", "Genes to label on volcano",
                 value=20L, min=0L, max=100L, step=5L),
    tags$div(class="text-muted mt-1", style="font-size:.78em;line-height:1.4;",
             icon("circle-info"),
             " Significance thresholds (\u03b1, log\u2082FC) from Parameters."),

    tags$hr(style="border-color:#ffffff30;margin:14px 0;"),

    tags$div(style="font-size:.73em;font-weight:700;text-transform:uppercase;
                    letter-spacing:.08em;color:#00F0BE;margin-bottom:8px;",
             icon("table"), " DEG Table Filter"),
    textInput("de_gene_filter", "Filter table by gene / protein ID",
              value="", placeholder="e.g. SERPINB3, CCL26\u2026"),
    tags$div(class="text-muted mt-1", style="font-size:.78em;line-height:1.4;",
             icon("circle-info"),
             " Partial match, case-insensitive. Use column headers in table for additional filters.")
)
```

## Row {height="540px"}

::: {.card title="Volcano Plot"}

```{r}
tags$p(style="font-size:.83em;color:#6c757d;margin-bottom:8px;",
       icon("circle-info"),
       " Red = significant (non-imputed); orange = significant (imputed); grey = not significant.",
       " LDV/LDV proteins have NA adj.p and are never labelled significant.")
plotlyOutput("de_volcano_plot", height="420px")
```

<p style="font-size:.82em;color:#6c757d;margin-top:6px;padding:0 2px;">
Dashed lines mark the adj.p (\u03b1) and |log\u2082FC| thresholds set in Parameters.
Hover any point to see gene name, protein ID, fold change, and adjusted p-value.
Use the <em>Effect type</em> selector to switch between all proteins, stable (common) effect, and interaction-driven proteins.
</p>
:::

## Row {height="530px"}

::: {.card title="Deregulogram — Factor-Dependent Effects"}

```{r}
uiOutput("de_dereg_panel_ui")
```

:::

## Row

::: {.card title="Differentially Expressed Proteins Table"}

```{r}
tags$p(style="font-size:.83em;color:#6c757d;margin-bottom:8px;",
       icon("circle-info"),
       " Showing significant proteins for the selected contrast and effect type.",
       " Use the Gene Filter (sidebar) or column-header search boxes to narrow results.",
       " Download CSV or Excel with the buttons above the table.")
DT::DTOutput("de_deg_table")
```

<p style="font-size:.82em;color:#6c757d;margin-top:8px;padding:0 2px;">
Proteins are sorted by adjusted p-value (ascending).
The <b>Imputed</b> flag indicates proteins with at least one LDV or mean-imputed value.
For interaction-effect results, log\u2082FC and adj.p refer to the interaction coefficient, not the main effect.
</p>
:::
